Summary Statistics

### 2025-03-13 extensive_tests/draft-sumstat-vignette.Rmd
added line breaks to make comparisons easier
Updating vignette with sumstat for mean number of significat freq changes per island and cherry

Intermediate values (e.g., obtained from pooled empirical data) should be 
categorized before analysis.

@@

```{r}
# 1 tip / sample / replicate
sample_n <- 1
index_islands <- c(1, 3)
index_nonislands <- c(2, 4)
data <- list(c(.5, .5, 0, 0, 0, .5), c(.5, 0, 0, .5), c(.5, .5, 0), c(0, 0, .5)) # tip 1
get_islandMeanFreqP(index_islands, data, sample_n)
get_nonislandMeanFreqP(index_nonislands, data, sample_n)
get_islandMeanFreqM(index_islands, data, sample_n)
get_nonislandMeanFreqM(index_nonislands, data, sample_n)
```

```{r}
# 2 tip / sample / replicate
sample_n <- 2
index_islands <- c(1, 3)
index_nonislands <- c(2, 4)
data <- list(
  list(c(.5, .5, 0, 0, 0, .5), c(.5, 0, 0, .5), c(.5, .5, 0), c(0, 0, .5)), # tip 1
  list(c(0, .5, .5, 1, 1, .5), c(1, .5, 1, .5), c(0, .5, .5), c(1, .5, 1)) # tip 2
  )
get_islandMeanFreqP(index_islands, data, sample_n)
get_nonislandMeanFreqP(index_nonislands, data, sample_n)
get_islandMeanFreqM(index_islands, data, sample_n)
get_nonislandMeanFreqM(index_nonislands, data, sample_n)
```
Note that in the calculation of the means the relative frequencies $x_{i, t}$, 
$y_{j, t}$, $w_{i, t}$ and $z_{j, t}$ are **not** weighted by the lengths of 
island $i$ or non-island $j$.

@@

```{r}
# 1 tip / sample / replicate
sample_n <- 1
index_islands <- c(1, 3)
index_nonislands <- c(2, 4)
data <- list(c(.5, .5, 0, 1, 1, .5), c(.5, 0, 1, .5), c(.5, .5, 0), c(0, 0, .5))
get_islandSDFreqP(index_islands, data, sample_n)
get_nonislandSDFreqP(index_nonislands, data, sample_n)
get_islandSDFreqM(index_islands, data, sample_n)
get_nonislandSDFreqM(index_nonislands, data, sample_n)
```

```{r}
# 2 tip / sample / replicate
sample_n <- 2
index_islands <- c(1, 3)
index_nonislands <- c(2, 4)
data <- list(
  list(c(.5, .5, 0, 0, 0, 1), c(.5, 0, 0, .5), c(1, .5, 0), c(0, 0, .5)), # tip 1
  list(c(0, .5, .5, 1, 1, .5), c(1, .5, 1, .5), c(0, .5, .5), c(1, .5, 1)) # tip 2
  )
get_islandSDFreqP(index_islands, data, sample_n)
get_nonislandSDFreqP(index_nonislands, data, sample_n)
get_islandSDFreqM(index_islands, data, sample_n)
get_nonislandSDFreqM(index_nonislands, data, sample_n)
```

@@

```{r}
# 1 tip / sample / replicate
sample_n <- 1
index_islands <- c(1, 3)
index_nonislands <- c(2, 4) 
data <- list(c(.5, 0, 0, 0, .5, .5, .5, .5, .5, 1, .5, 0, 0, 0, .5, .5, .5, .5, 
               .5, 1, .5, 0, 0, 0, .5, .5, .5, .5, .5, 1), # 30 sites
               c(.5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, 
                 .5, .5, 1, 1, 1, .5), # 25 sites
               c(.5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, 
                 .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 
                 0, 0, .5, 1), # 40 sites
               c(1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, .5, .5, .5, 0, 0, 0, 
                 .5, 1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 
                 .5, 0, 0, 0, .5)) # 35 sites
compute_meanCor_i(index_islands, minN_CpG = 10, 
                  shore_length = 5, data, sample_n = 1)
compute_meanCor_ni(index_nonislands, minN_CpG = 10, 
                   shore_length = 5, data, sample_n = 1)
```

```{r}
# 2 tip / sample / replicate
sample_n <- 2
index_islands <- c(1, 3)
index_nonislands <- c(2, 4)
data <- list(
  # tip 1
    list(c(.5, 0, 0, 0, .5, .5, .5, .5, .5, 1, .5, 0, 0, 0, .5, .5, .5, .5, .5, 
           1, .5, 0, 0, 0, .5, .5, .5, .5, .5, 1), # 30 sites
         c(.5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5, 
           .5, 1, 1, 1, .5), # 25 sites
         c(.5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, 
           .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, 
           .5, 1), # 40 sites
         c(1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 
           1, 1, 1, .5, .5, .5, 0, 0, 0, .5, .5, 0, 0, 0, .5)), # 35 sites
  # tip 2
    list(c(.5, 0, 0, .5, .5, .5, 0, 0, .5, 1, .5, 0, 0, 0, 0, .5, .5, 1, 1, 1, 
           .5, 0, 0, 0, .5, .5, 1, 1, 1, 1), # 30 sites
         c(.5, .5, 1, 1, .5, .5, 1, 1, 1, .5, .5, 0, 0, 0, .5, .5, 1, 1, 1, .5, 
           .5, 1, 1, 1, .5), # 25 sites
         c(.5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, .5, .5, 0, 0, .5, 1, 
           1, 1, 1, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, 
           .5, 1), # 40 sites
         c(1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, 1, .5, .5, 0, 0, 0, .5, 1, 
           1, 1, .5, .5, .5, .5, .5, 0, .5, .5, .5, .5, 0, .5)) # 35 sites
  )
compute_meanCor_i(index_islands, minN_CpG = 10, 
                  shore_length = 5, data, sample_n = 2)
compute_meanCor_ni(index_nonislands, minN_CpG = 10, 
                   shore_length = 5, data, sample_n = 2)
```

@@

MeanSiteFChange_cherry(data = data, 
                       tree = tree, 
                       index_islands = index_islands, 
                       index_nonislands = index_nonislands)
```

@@

>>>>

Intermediate values (e.g., obtained from pooled empirical data) should be 
categorized before analysis with `categorize_siteMethSt`.

```{r}
non_categorized_data <- list(
  # Tip 1
    list(c(0.1, 0.7, 0.9), # First structure
         c(0.3, 0.5, 0.9)), # Second structure
  # Tip 2
    list(c(0.2, 0.8, 0.6), # First structure
         c(0.9, 0.4, 0.7)) # Second structure
  )

# Transform the data with custom thresholds
categorized_data <- categorize_siteMethSt(data, u_threshold = 0.15, m_threshold = 0.85)

categorized_data
```

@@

```{r}
# 1 tip / sample / replicate
sample_n <- 1
index_islands <- c(1, 3)
index_nonislands <- c(2, 4)
data <- list(c(.5, .5, 0, 0, 0, .5), c(.5, 0, 0, .5), c(.5, .5, 0), c(0, 1, .5)) # tip 1
get_islandMeanFreqP(index_islands, data, categorized_data = T, sample_n)
get_nonislandMeanFreqP(index_nonislands, data, categorized_data = T, sample_n)
get_islandMeanFreqM(index_islands, data, categorized_data = T, sample_n)
get_nonislandMeanFreqM(index_nonislands, data, categorized_data = T, sample_n)
```

```{r}
# 2 tip / sample / replicate
sample_n <- 2
index_islands <- c(1, 3)
index_nonislands <- c(2, 4)
data <- list(
  list(c(.5, .5, 0, 0, 0, .5), c(.5, 0, 0, .5), c(.5, .5, 0), c(0, 0, .5)), # tip 1
  list(c(0, .5, .5, 1, 1, .5), c(1, .5, 1, .5), c(0, .5, .5), c(1, .5, 1)) # tip 2
  )
get_islandMeanFreqP(index_islands, data, categorized_data = T, sample_n)
get_nonislandMeanFreqP(index_nonislands, data, categorized_data = T, sample_n)
get_islandMeanFreqM(index_islands, data, categorized_data = T, sample_n)
get_nonislandMeanFreqM(index_nonislands, data, categorized_data = T, sample_n)
```
Note that in the calculation of the means the relative frequencies $x_{i, t}$, 
$y_{j, t}$, $w_{i, t}$ and $z_{j, t}$ are **not** weighted by the lengths of 
island $i$ or non-island $j$.

@@

```{r}
# 1 tip / sample / replicate
sample_n <- 1
index_islands <- c(1, 3)
index_nonislands <- c(2, 4)
data <- list(c(.5, .5, 0, 1, 1, .5), c(.5, 0, 1, .5), c(.5, .5, 0), c(0, 0, .5))
get_islandSDFreqP(index_islands, data, categorized_data = T, sample_n)
get_nonislandSDFreqP(index_nonislands, data, categorized_data = T, sample_n)
get_islandSDFreqM(index_islands, data, categorized_data = T, sample_n)
get_nonislandSDFreqM(index_nonislands, data, categorized_data = T, sample_n)
```

```{r}
# 2 tip / sample / replicate
sample_n <- 2
index_islands <- c(1, 3)
index_nonislands <- c(2, 4)
data <- list(
  list(c(.5, .5, 0, 0, 0, 1), c(.5, 0, 0, .5), c(1, .5, 0), c(0, 0, .5)), # tip 1
  list(c(0, .5, .5, 1, 1, .5), c(1, .5, 1, .5), c(0, .5, .5), c(1, .5, 1)) # tip 2
  )
get_islandSDFreqP(index_islands, data, categorized_data = T, sample_n)
get_nonislandSDFreqP(index_nonislands, data, categorized_data = T, sample_n)
get_islandSDFreqM(index_islands, data, categorized_data = T, sample_n)
get_nonislandSDFreqM(index_nonislands, data, categorized_data = T, sample_n)
```

@@

```{r}
# 1 tip / sample / replicate
sample_n <- 1
index_islands <- c(1, 3)
index_nonislands <- c(2, 4) 
data <- list(c(.5, 0, 0, 0, .5, .5, .5, .5, .5, 1, .5, 0, 0, 0, .5, .5, .5, .5, 
               .5, 1, .5, 0, 0, 0, .5, .5, .5, .5, .5, 1), # 30 sites
               c(.5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, 
                 .5, .5, 1, 1, 1, .5), # 25 sites
               c(.5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, 
                 .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 
                 0, 0, .5, 1), # 40 sites
               c(1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, .5, .5, .5, 0, 0, 0, 
                 .5, 1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 
                 .5, 0, 0, 0, .5)) # 35 sites
compute_meanCor_i(index_islands, minN_CpG = 10, 
                  shore_length = 5, data, sample_n = 1, categorized_data = T)
compute_meanCor_ni(index_nonislands, minN_CpG = 10, 
                   shore_length = 5, data, sample_n = 1, categorized_data = T)
```

```{r}
# 2 tip / sample / replicate
sample_n <- 2
index_islands <- c(1, 3)
index_nonislands <- c(2, 4)
data <- list(
  # tip 1
    list(c(.5, 0, 0, 0, .5, .5, .5, .5, .5, 1, .5, 0, 0, 0, .5, .5, .5, .5, .5, 
           1, .5, 0, 0, 0, .5, .5, .5, .5, .5, 1), # 30 sites
         c(.5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5, .5, 1, 1, 1, .5, 
           .5, 1, 1, 1, .5), # 25 sites
         c(.5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, 
           .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, 
           .5, 1), # 40 sites
         c(1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 
           1, 1, 1, .5, .5, .5, 0, 0, 0, .5, .5, 0, 0, 0, .5)), # 35 sites
  # tip 2
    list(c(.5, 0, 0, .5, .5, .5, 0, 0, .5, 1, .5, 0, 0, 0, 0, .5, .5, 1, 1, 1, 
           .5, 0, 0, 0, .5, .5, 1, 1, 1, 1), # 30 sites
         c(.5, .5, 1, 1, .5, .5, 1, 1, 1, .5, .5, 0, 0, 0, .5, .5, 1, 1, 1, .5, 
           .5, 1, 1, 1, .5), # 25 sites
         c(.5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, .5, .5, 0, 0, .5, 1, 
           1, 1, 1, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, .5, 1, .5, 0, 0, 
           .5, 1), # 40 sites
         c(1, 1, 1, .5, .5, .5, 0, 0, 0, .5, 1, 1, 1, 1, .5, .5, 0, 0, 0, .5, 1, 
           1, 1, .5, .5, .5, .5, .5, 0, .5, .5, .5, .5, 0, .5)) # 35 sites
  )
compute_meanCor_i(index_islands, minN_CpG = 10, 
                  shore_length = 5, data, sample_n = 2, categorized_data = T)
compute_meanCor_ni(index_nonislands, minN_CpG = 10, 
                   shore_length = 5, data, sample_n = 2, categorized_data = T)
```

@@

MeanSiteFChange_cherry(data = data, 
                       categorized_data = T, 
                       tree = tree, 
                       index_islands = index_islands, 
                       index_nonislands = index_nonislands)
```

@@

- `compare_CherryFreqs(tip1, tip2)` to perform a chi-squared test to compare the
distribution of methylation states between two cherry tips.

- `pValue_CherryFreqsChange_i(data, index_islands, tree)` uses `compare_CherryFreqs(tip1, tip2)` to get for each cherry 
and each island the chi-square p-value of the comparison of the distribution of
methylation states between two cherry tips.

- `mean_CherryFreqsChange_i(data, categorized_data = FALSE, index_islands, tree, pValue_threshold)`. The function
uses `pValue_CherryFreqsChange_i(data, index_islands, tree)` and counts the mean
number of significant changes per island at each cherry.

```{r}
# Set example tree and methylation data
  tree <- "((a:1,b:1):2,c:2);"
  data <- list(
    #Tip a
    list(c(rep(1,9), rep(0,1)), # Structure 1: island
         c(rep(0,9), 1), # Structure 2: non-island
         c(rep(0,9), rep(0.5,1))),  # Structure 3: island
    #Tip b
    list(c(rep(0,9), rep(0.5,1)),  # Structure 1: island
         c(rep(0.5,9), 1), # Structure 2: non-island
         c(rep(0,9), rep(0,1))), # Structure 3: island
    #Tip c
    list(c(rep(1,9), rep(0.5,1)),  # Structure 1: island
         c(rep(0.5,9), 1), # Structure 2: non-island
         c(rep(0,9), rep(0.5,1)))) # Structure 3: island

  mean_CherryFreqsChange_i(data, categorized_data = T,
                           index_islands, tree,
                           pValue_threshold = 0.05)
```
